Phase Change Memory (PCM) is a new \emph{non-volatile} memory technology
that is comparable to traditional DRAM with regard to read latency,
and markedly superior with regard to storage density and idle
power consumption. Due to these desirable characteristics, PCM is
expected to play a significant role in the next generation of computing
systems. However, it also has limitations in the form of expensive writes
and limited write endurance. Accordingly, recent research has investigated
how database engines may be redesigned to suit DBMS deployments on the
new technology.

In this paper, we address the pragmatic goal of minimally altering current
implementations of database operators to make them PCM-conscious,
the objective being to facilitate an easy transition to the new
technology. Specifically, we target the implementations of the
``workhorse'' database operators: \emph{sort}, \emph{hash join} and
\emph{group-by}, and rework them to substantively improve the write
performance without compromising on execution times. Concurrently, we
provide simple but effective \emph{estimators} of the writes incurred
by the new techniques, and these estimators are leveraged for 
integration with the query optimizer.

Our new techniques are evaluated on TPC-H benchmark queries with
regard to the following metrics: number of writes, response times and wear distribution. The experimental results indicate that the PCM-conscious
operators collectively reduce the number of writes by a factor of 2 to
3, while concurrently improving the query response times by about 20\% to
30\%.  When combined with the appropriate plan choices, the improvements
are even higher.  In essence, our algorithms provide both short-term and
long-term benefits.  These outcomes augur well for database engines that
wish to leverage the impending transition to PCM-based computing.
