\section{Conclusion}
\label{conclusion}
Designing database query execution algorithms for PCM platforms requires a
change in perspective from the traditional assumptions of symmetric read
and write overheads.  We presented here a variety of minimally modified algorithms
for the workhorse database operators: \emph{sort}, \emph{hash join} and
\emph{group-by}, which were constructed with a view towards simultaneously
reducing both the number of writes and the response time. Through detailed
experimentation on complete TPC-H benchmark queries, we showed that
substantial improvements on these metrics can be obtained as compared
to their contemporary PCM-oblivious counterparts.  Collaterally, the
PCM cell lifetimes are also greatly extended by the new approaches.

Using our write estimators for uniformly distributed data, we presented
a redesigned database optimizer, thereby incorporating PCM-consciousness
in all layers of the database engine. We also presented initial
results showing how this can influence plan choices, and improve the write
performance by a substantial margin.  While our experiments were conducted
on a PCM simulator, the cycle-accurate nature of the simulator makes it
likely that similar performance will be exhibited in the real world as
well. In our future work, we would like to design write estimators that
leverage the metadata statistics to accurately predict writes for skewed
data. Additionally, we wish to design multi-objective
optimization algorithms for query plan selection with provable performance
guarantees.

Overall, the results of this paper augur well for an easy migration of
current database engines to leverage the benefits of tomorrow's PCM-based
computing platforms.
